The Proofs

### 2023-10-17 proofs/main.tex
Fix structure
Restructure, add biblio
Restructure
WIP
wip

\UseRawInputEncoding
% vim: ft=tex
\documentclass[a4,natbib=false]{article}
\usepackage[a4paper, total={8in, 10in}]{geometry}
\usepackage{hyperref}
\usepackage{mathpartir}

\input{prelude.tex}

\newcommand{\ottDir}{../ott/_gen}
\newcommand{\genDir}{_gen}
\input{\ottDir/included.tex}

% \renewcommand{\ottdruleEOneNVarName}[0]{ Hello \def\@currentlabelname{FOO} \phantomsection  }
% \newcommand{\ottdrulename}[1]{\textsc{#1}}

% \renewcommand{\ottdrule}[4][]{{\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}%
%   }\ruleLabel{#4}{#4}}

% \renewcommand{\ottdrule}[4][]{%
%   {\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}}%
%   \mpr@label{\textsc{#1}}%
% }
% \DeclareDocumentCommand \redrule {m m o}{%
%   \inferrule*[vcenter,left={#3:}]{}{#1 \tred #2}
%   \mpr@label{\textsc{#3}}
% }

% \renewcommand{\ottdrulename}[2][#1]{\textsc{#1} \label{}}

% \renewcommand{\ottdruleEOneNVarName}[0]{foo }

% \makeatletter
% \renewcommand{\ottdruleEOneNVar}[1]{\ottdrule[#1]{%
%   }{
%     % \protected@edef\@currentlabelname{foo}
%     \def\@currentlabelname{foo}%
%     \phantomsection
%     \label{boo}
%     \alpha ^{-}   \eqEOne   \alpha ^{-}
%   }{%
%     {\ottdruleEOneNVarName}{}%
%   } }
% \makeatother

\newcommand{\depth}[1]{\ensuremath{\mathsf{depth}(#1)}}
\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}

\newcommand{\ruleref}[1]{Rule \nameref{#1}}

% ord varset in uN = varset'

\renewcommand{\ottdruleONVarInName}[0]{(Var$_{-\in}^{\text{Ord}}$)}
\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{-\notin}^{\text{Ord}}$)}
\renewcommand{\ottdruleONUVarName}[0]{(UVar$_{-} ^{\text{Ord}}$)}
\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow^{\text{Ord}}$)}
\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow^{\text{Ord}}$)}
\renewcommand{\ottdruleOForallName}[0]{($\forall^{\text{Ord}}$)}

% ord varset in uP = varset'

\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{+\in}^{\text{Ord}}$)}
\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{+\notin}^{\text{Ord}}$)}
\renewcommand{\ottdruleOPUVarName}[0]{(UVar$_{+}^{\text{Ord}}$)}
\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow^{\text{Ord}}$)}
\renewcommand{\ottdruleOExistsName}[0]{($\exists^{\text{Ord}}$)}

% nf(N) = M
\renewcommand{\ottdruleNrmNVarName}[0]{(Var$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow^{\text{nf}}$)}
\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow^{\text{nf}}$)}
\renewcommand{\ottdruleNrmForallName}[0]{($\forall^{\text{nf}}$)}

% nf(P) = Q
\renewcommand{\ottdruleNrmPVarName}[0]{(Var$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow^{\text{nf}}$)}
\renewcommand{\ottdruleNrmExistsName}[0]{($\exists^{\text{nf}}$)}

% N ≈ M

\renewcommand{\ottdruleEOneNVarName}[0]{(Var$_-^{\eqEOne}$)}
\renewcommand{\ottdruleEOneShiftUName}[0]{($\uparrow^{\eqEOne}$)}
\renewcommand{\ottdruleEOneArrowName}[0]{($\rightarrow^{\eqEOne}$)}
\renewcommand{\ottdruleEOneForallName}[0]{($\forall^{\eqEOne}$)}

% P ≈ Q
\renewcommand{\ottdruleEOnePVarName}[0]{(Var$_+^{\eqEOne}$)}
\renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
\renewcommand{\ottdruleEOneExistsName}[0]{($\exists^{\eqEOne}$)}

% G ⊢ N ≤1 M

\renewcommand{\ottdruleDOneNVarName}[0]{(Var$_-^{\subDOne}$)}
\renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
\renewcommand{\ottdruleDOneArrowName}[0]{($\rightarrow^{\subDOne}$)}
\renewcommand{\ottdruleDOneForallName}[0]{($\forall^{\subDOne}$)}

% G ⊢ P ≥1 Q
\renewcommand{\ottdruleDOnePVarName}[0]{(Var$_+^{\supDOne}$)}
\renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
\renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}

% G ⊢ N ≈1 M
\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne_{-}$)}

% G ⊢ P ≈1 Q
\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne_{+}$)}

% G ⊨ iP1 ∨ iP2 = iQ
\renewcommand{\ottdruleLUBVarName}[0]{(Var$^{\vee}$)}
\renewcommand{\ottdruleLUBShiftName}[0]{($\downarrow^{\vee}$)}
\renewcommand{\ottdruleLUBExistsName}[0]{($\exists^{\vee}$)}
\renewcommand{\ottdruleLUBUpgradeName}[0]{(Upg)}

% G ; Θ ⊨ uN ≤ iM ⫤ us
\renewcommand{\ottdruleANVarName}[0]{(Var$_-^{\subA}$)}
\renewcommand{\ottdruleAShiftUName}[0]{($\uparrow^{\subA}$)}
\renewcommand{\ottdruleAArrowName}[0]{($\rightarrow^{\subA}$)}
\renewcommand{\ottdruleAForallName}[0]{($\forall^{\subA}$)}

% G ; Θ ⊨ iP ≥ uQ ⫤ us
\renewcommand{\ottdruleAPVarName}[0]{(Var$_+^{\supA}$)}
\renewcommand{\ottdruleAShiftDName}[0]{($\downarrow^{\supA}$)}
\renewcommand{\ottdruleAExistsName}[0]{($\exists^{\supA}$)}
\renewcommand{\ottdruleAPUVarName}[0]{(UVar$^{\supA}$)}

% Γ ⊢ scE1 & scE2 = scE3 :: :: E :: 'E'
\renewcommand{\ottdruleSCMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
\renewcommand{\ottdruleSCMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
\renewcommand{\ottdruleSCMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
\renewcommand{\ottdruleSCMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
\renewcommand{\ottdruleSCMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}

% Γ ; Θ ⊨ uN ≈u iM ⫤ UC 
\renewcommand{\ottdruleUNVarName}[0]{(Var$_{-}^{[[≈u]]}$)}
\renewcommand{\ottdruleUShiftUName}[0]{($\uparrow^{[[≈u]]}$)}
\renewcommand{\ottdruleUArrowName}[0]{($\rightarrow^{[[≈u]]}$)}
\renewcommand{\ottdruleUForallName}[0]{($\forall^{[[≈u]]}$)}
\renewcommand{\ottdruleUNUVarName}[0]{(UVar$_{-}^{[[≈u]]}$)}

% Γ ; Θ ⊨ uP ≈u iQ ⫤ UC 
\renewcommand{\ottdruleUPVarName}[0]{(Var$_{+}^{[[≈u]]}$)}
\renewcommand{\ottdruleUShiftDName}[0]{($\downarrow^{[[≈u]]}$)}
\renewcommand{\ottdruleUExistsName}[0]{($\exists^{[[≈u]]}$)}
\renewcommand{\ottdruleUPUVarName}[0]{(UVar$_{+}^{[[≈u]]}$)}

% G ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )
\renewcommand{\ottdruleAUPVarName}[0]{(Var$_{+}^{[[≈au]]}$)}
\renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
\renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}

% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
\renewcommand{\ottdruleAUNVarName}[0]{(Var$_{-}^{[[≈au]]}$)}
\renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
\renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
\renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
\renewcommand{\ottdruleAUAUName}[0]{(AU)}

% Γ ⊢ iN 
\renewcommand{\ottdruleWFTNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
\renewcommand{\ottdruleWFTShiftUName}[0]{($\uparrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFTArrowName}[0]{($\rightarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFTForallName}[0]{($\forall^{\text{WF}}$)}

% Γ ⊢ iP
\renewcommand{\ottdruleWFTPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
\renewcommand{\ottdruleWFTShiftDName}[0]{($\downarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFTExistsName}[0]{($\exists^{\text{WF}}$)}

% Γ ; Ξ ⊢ uN
\renewcommand{\ottdruleWFATNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATNUVarName}[0]{(UVar$_{-}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATShiftUName}[0]{($\uparrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFATArrowName}[0]{($\rightarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFATForallName}[0]{($\forall^{\text{WF}}$)}

% Γ ; Ξ ⊢ uP
\renewcommand{\ottdruleWFATPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATPUVarName}[0]{(UVar$_{+}^{\text{WF}}$)}
\renewcommand{\ottdruleWFATShiftDName}[0]{($\downarrow^{\text{WF}}$)}
\renewcommand{\ottdruleWFATExistsName}[0]{($\exists^{\text{WF}}$)}

% Γ ⊢ iP : scE
\renewcommand{\ottdruleSATSCESupName}[0]{($[[:≥]]_{+}^\text{sat}$)}
\renewcommand{\ottdruleSATSCEPEqName}[0]{($[[:≈]]_{+}^\text{sat}$)}

% Γ ⊢ iN : scE
\renewcommand{\ottdruleSATSCENEqName}[0]{($[[:≈]]_{-}^\text{sat}$)}

% Γ ; Φ ⊢ v : iP 
\renewcommand{\ottdruleDTVarName}[0]{(Var$^{\text{inf}}$)}
\renewcommand{\ottdruleDTThunkName}[0]{($\{\}^{\text{inf}}$)}
\renewcommand{\ottdruleDTPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
\renewcommand{\ottdruleDTPEquivName}[0]{($[[≈]]_+^{\text{inf}}$)}

% Γ ; Φ ⊢ c : iN 
\renewcommand{\ottdruleDTtLamName}[0]{($\lambda^{\text{inf}}$)}
\renewcommand{\ottdruleDTTLamName}[0]{($\Lambda^{\text{inf}}$)}
\renewcommand{\ottdruleDTReturnName}[0]{(ret$^{\text{inf}}$)}
\renewcommand{\ottdruleDTVarLetName}[0]{(let$^{\text{inf}}$)}
\renewcommand{\ottdruleDTAppLetName}[0]{(let$_@^{\text{inf}}$)}
\renewcommand{\ottdruleDTAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
\renewcommand{\ottdruleDTUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
\renewcommand{\ottdruleDTNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
\renewcommand{\ottdruleDTNEquivName}[0]{($[[≈]]_-^{\text{inf}}$)}

% Γ ; Φ ⊢ iN ● args ⇒> iM 
\renewcommand{\ottdruleDTEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleDTArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleDTForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}

% Γ ; Φ ⊨ v : iP  
\renewcommand{\ottdruleATVarName}[0]{(Var$^{\text{inf}}$)}
\renewcommand{\ottdruleATThunkName}[0]{($\{\}^{\text{inf}}$)}
\renewcommand{\ottdruleATPAnnotName}[0]{(ann$_+^{\text{inf}}$)}

% Γ ; Φ ⊨ c : iN 
\renewcommand{\ottdruleATtLamName}[0]{($\lambda^{\text{inf}}$)}
\renewcommand{\ottdruleATTLamName}[0]{($\Lambda^{\text{inf}}$)}
\renewcommand{\ottdruleATReturnName}[0]{(ret$^{\text{inf}}$)}
\renewcommand{\ottdruleATVarLetName}[0]{(let$^{\text{inf}}$)}
\renewcommand{\ottdruleATAppLetName}[0]{(let$_@^{\text{inf}}$)}
\renewcommand{\ottdruleATAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
\renewcommand{\ottdruleATUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
\renewcommand{\ottdruleATNAnnotName}[0]{(ann$_-^{\text{inf}}$)}

% Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC 
\renewcommand{\ottdruleATEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleATArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
\renewcommand{\ottdruleATForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}

% scE1 singular with iP
\renewcommand{\ottdruleSINGPEqName}[0]{($[[≈]]_{+}^{\text{sing}}$)}
\renewcommand{\ottdruleSINGSupVarName}[0]{($[[:≥]]\alpha^{\text{sing}}$)}
\renewcommand{\ottdruleSINGSupShiftName}[0]{($[[:≥]][[↓]]^{\text{sing}}$)}

% scE1 singular with iN
\renewcommand{\ottdruleSINGNEqName}[0]{($[[≈]]_{-}^{\text{sing}}$)}

@@

@@

\begin{algorithm}[Substitution Application]
  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
  It is defined naturally as follows:
    \begin{multicols}{2}
      \begin{itemize}
        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
          that is the variable capture never happens);
      \end{itemize}
    \end{multicols}
\end{algorithm}

@@

  \begin{multicols}{2}
    \ottdefnDOneNsub{}
    \columnbreak\\
    \ottdefnDOnePsup{}
  \end{multicols}

  \begin{multicols}{2}
    \ottdefnDOneNeq{}
    \columnbreak\\
    \ottdefnDOnePeq{}
  \end{multicols}
\end{definition}

@@

In the algorithmic system, we extend the grammar of types
by adding positive and negative \emph{algorithmic variables}
($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
They represent the unknown types, which will be inferred by the algorithm.
This way, we add two base cases to the grammar of 
positive and negative types, and use highlight to denote that the type
can potentially contain algorithmic variables.

@@

In several places of our algorithm, in particular, during
algorithmic subtyping,
we turn a declarative type into the algorithmic one
via replacing certain type variables with fresh algorithmic variables.
We call this procedure \emph{variable algorithmization}, and define it as follows.

@@

  \begin{multicols}{2}
  \ottdefnONVar{}
  \columnbreak\\
  \ottdefnOPVar{}
  \end{multicols}

@@

  Analogously,  we define normalization of algorithmic types 
  by adding base cases:

@@

\begin{figure}[h]
  \centering
  \begin{tikzpicture}
    [>={Stealth[scale=2]},node distance=2.6cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]

    % Define nodes
    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
    \node[left=2cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
    \node[right=2cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
    \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
    \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};

@@

Now, we present the subtyping algorithm itself.
Although the algorithm is presented as a single procedure,
is important for the structure of the proof that the positive subtyping algorithm
does not invoke the negative one. This way, the correctness of the positive 
subtyping will be proved independently, and used afterwards to prove the
correctness of the negative subtyping.

\begin{algorithm}[Subtyping]
  \label{alg:subtyping}
  \hfill

  \begin{multicols}{2}
  \ottdefnANsub{}
  \columnbreak\\
  \ottdefnAPsup{}
  \end{multicols}
\end{algorithm}

@@

Another dual pair of constructions are 
$[[return v]]$ and $[[{c}]]$. 
The former allows us to embed a value in a pure 
computations. The latter, on the contrary,
encapsulates a thunk of computation in a value.

Finally, the language has a number standard constructions:
lambda-abstractions $[[λx:iP.c]]$,
standard let-bindings $[[let x = v; c]]$,
and type annotations that can be added to any value or computation:
$[[(v:iP)]]$ and $[[(c:iN)]]$.

@@

Next, we define the specification of the type 
inference for our language. First, we introduce 
variable context specifying the types of the variables 
in the scope of the current rule.

@@

The specification is represented by an inference system of
three mutually recursive judgments:
positive inference $[[Γ ; Φ ⊢ v : iP]]$,
negative type inference $[[Γ ; Φ ⊢ c : iN]]$, and
application type inference $[[Γ ; Φ ⊢ iN ● args ⇒> iM ]]$.
In the premises, the inference rules also refer to 
the declarative subtyping (\cref{def:subDOne}),
type well-formedness (\cref{alg:wf}), 
and normalization (\cref{alg:type-nf}).
\begin{itemize}
  \item $[[Γ ; Φ ⊢ v : iP]]$ (and symmetrically, $[[Γ ; Φ ⊢ c : iN]]$)
    means that under the type context $[[Γ]]$ and
    the variable context $[[Φ]]$, for the value $[[v]]$,
    type $[[iP]]$ is inferrable. It guarantees that 
    $[[v]]$ is well-formed in $[[Γ]]$ and $[[Φ]]$ in
    the standard sense.
  \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
    judgement. It means that if a head of type $[[iN]]$ 
    is applied to list of values $[[args]]$, 
    then the resulting computation can be typed as $[[iM]]$.
\end{itemize}

@@

Let us discuss selected rules of the declarative system:
\begin{itemize}
  \item \ruleref{\ottdruleDTVarLabel}
    says that the type of a variable is inferred from the context.
  \item \ruleref{\ottdruleDTThunkLabel} says
    that the type of a thunk is inferred by shifting up the type of the 
    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
    infers the type of a return by shifting down the type of the
    contained value.
  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
    They allow the inferred type to be refined by annotating it with a supertype.
  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
    mean that th declarative system allows to infer any type from the equivalence class.
  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
    and its first premise inferring the existential type of the value being unpacked.
    It is important however that the inferred existential type is normalized. 
    This is because there might be multiple equivalent existential types 
    with different order or even number of quantified variables, 
    and to bind them, the algorithm needs to fix the canonical one.
  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
    with annotated let-bindings. The first premise infers the type of the head of the application,
    which must be a thunked computation. Then if after applying it 
    to tha arguments, the resulting type can be equated to the annotated one,
    we infer the body of the let-binding in the context extended with the bound variable.
  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
    but it is used when the type of the application is unique, and thus, the annotation
    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
\end{itemize}

@@

\begin{algorithm}
  \hfill
  \begin{multicols}{2}
  \ottdefnATNInf{}

  \columnbreak
  \ottdefnATPInf{}
  \ottdefnATSpinInf{}
  \end{multicols}
\end{algorithm}

@@

The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
of the algorithmic typing check wether the constraint $[[SC]]$
is uniquely defines a substitution satisfying it, and if it does
returns such a substitution as the result.
To do that, we define a partial function $[[SC singular with uσ]]$,
taking a subtyping constraint $[[SC]]$ and returning a substitution 
$[[uσ]]$---the only possible solution of $[[SC]]$.

>>>>

\PassOptionsToPackage{prologue,dvipsnames}{xcolor}

% \UseRawInputEncoding
% vim: ft=tex

% \documentclass[acmsmall,natbib=false]{article}
% \usepackage[a4paper, total={8in, 10in}]{geometry}

\documentclass[acmsmall,natbib=false,review,anonymous]{acmart}
\input{../prelude.tex}

\newcommand{\ottDir}{../ott/_gen}
\newcommand{\genDir}{_gen}
\input{\ottDir/defs.tex}
\input{\ottDir/renew-ott.tex}

\usepackage{../bibliography}
% \usepackage{amssymb}

\usepackage[matha]{mathabx}

% https://tex.stackexchange.com/questions/85033/colored-symbols/85035#85035
% \newcommand*{\mathcolor}{}
% \def\mathcolor#1#{ \mathcoloraux{#1} }
% \newcommand*{\mathcoloraux}[3]{%
%   \protect\leavevmode
%   \begingroup
%   \color#1{#2}#3%
%   \endgroup
% }

@@

\title{The Proofs}

\maketitle

@@

\begin{algorithm}[Substitution Application]
  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
  It is defined naturally as follows:
    \begin{multicols}{2}
      \begin{itemize}
        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$
        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$
        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$
        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$
        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$
        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$ 
        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (assuming the variable capture never happens)
      \end{itemize}
    \end{multicols}
\end{algorithm}

@@

  \begin{multicols}{2}
    \ottdefnDOneNsub{}

    \ottdefnDOnePsup{}
  \end{multicols}
  \hfill

  \begin{multicols}{2}
    \ottdefnDOneNeq{}

    \ottdefnDOnePeq{}
  \end{multicols}
\end{definition}

@@

In the algorithmic system, we extend the grammar of types
by adding positive and negative \emph{algorithmic variables}
($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
They represent the unknown types, which will be inferred by the algorithm.
This way, we add two base cases to the grammar of 
positive and negative types and use highlight to denote that the type
can potentially contain algorithmic variables.

@@

In several places of our algorithm, in particular, during
algorithmic subtyping,
we turn a declarative type into an algorithmic one
via replacing certain type variables with fresh algorithmic variables.
We call this procedure \emph{variable algorithmization}, and define it as follows.

@@

  \ottdefnONVar{}
  \ottdefnOPVar{}

@@

  Analogously, we define the normalization of algorithmic types by adding base cases:

@@

\begin{figure}[h]
  \centering
  \begin{tikzpicture}
    [>={Stealth[scale=2]},node distance=2.4cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]

    % Define nodes
    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
    \node[left=1cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
    \node[right=1.4cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
    \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
    \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};

@@

Now, we present the subtyping algorithm itself.
Although the algorithm is presented as a single procedure,
is important for the structure of the proof that the positive subtyping algorithm
does not invoke the negative one. This way, the correctness of the positive 
subtyping will be proved independently and used afterwards to prove the
correctness of the negative subtyping.

\begin{algorithm}[Subtyping]
  \label{alg:subtyping}
  \hfill\\
  \ottdefnANsub{}
  \ottdefnAPsup{}
\end{algorithm}

@@

Another dual pair of constructions are 
$[[return v]]$ and $[[{c}]]$. 
The former allows us to embed a value in pure 
computations. The latter, on the contrary,
encapsulates a thunk of computation in a value.

Finally, the language has several standard constructions:
lambda-abstractions $[[λx:iP.c]]$,
standard let-bindings $[[let x = v; c]]$,
and type annotations that can be added to any value or computation:
$[[(v:iP)]]$ and $[[(c:iN)]]$.

@@

Next, we define the specification of the type 
inference for our language. First, we introduce 
variable context specifying the types of variables 
in the scope of the current rule.

@@

The specification is represented by an inference system of
three mutually recursive judgments:
positive inference $[[Γ ; Φ ⊢ v : iP]]$,
negative type inference $[[Γ ; Φ ⊢ c : iN]]$, and
application type inference $[[Γ ; Φ ⊢ iN ● args ⇒> iM ]]$.
In the premises, the inference rules also refer to 
the declarative subtyping (\cref{def:subDOne}),
type well-formedness (\cref{alg:wf}), 
and normalization (\cref{alg:type-nf}).
\begin{itemize}
  \item $[[Γ ; Φ ⊢ v : iP]]$ (and symmetrically, $[[Γ ; Φ ⊢ c : iN]]$)
    means that under the type context $[[Γ]]$ and
    the variable context $[[Φ]]$, for the value $[[v]]$,
    type $[[iP]]$ is inferrable. It guarantees that 
    $[[v]]$ is well-formed in $[[Γ]]$ and $[[Φ]]$ in
    the standard sense.
  \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
    judgment. It means that if a head of type $[[iN]]$ 
    is applied to list of values $[[args]]$, 
    then the resulting computation can be typed as $[[iM]]$.
\end{itemize}

@@

Let us discuss selected rules of the declarative system:
\begin{itemize}
  \item \ruleref{\ottdruleDTVarLabel}
    says that the type of a variable is inferred from the context.
  \item \ruleref{\ottdruleDTThunkLabel} says
    that the type of a thunk is inferred by shifting up the type of the 
    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
    infers the type of a return by shifting down the type of the
    contained value.
  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
    They allow the inferred type to be refined by annotating it with a supertype.
  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
    mean that th declarative system allows to infer any type from the equivalence class.
  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
    and its first premise inferring the existential type of the value being unpacked.
    It is important however that the inferred existential type is normalized. 
    This is because there might be multiple equivalent existential types 
    with different order or even number of quantified variables, 
    and to bind them, the algorithm needs to fix the canonical one.
  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
    with annotated let-bindings. The first premise infers the type of the head of the application,
    which must be a thunked computation. Then if after applying it 
    to the arguments, the resulting type can be equated to the annotated one,
    we infer the body of the let-binding in the context extended with the bound variable.
  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
    but it is used when the type of the application is unique, and thus, the annotation
    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
\end{itemize}

@@

\begin{algorithm}
  \hfill\\
  \ottdefnATNInf{}
  \ottdefnATPInf{}
  \ottdefnATSpinInf{}
\end{algorithm}

@@

The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
of the algorithmic typing check whether the constraint $[[SC]]$
is uniquely defines a substitution satisfying it, and if it does
returns such a substitution as the result.
To do that, we define a partial function $[[SC singular with uσ]]$,
taking a subtyping constraint $[[SC]]$ and returning a substitution 
$[[uσ]]$---the only possible solution of $[[SC]]$.
